Detecting Hate Speech with GPT-3

### 2022-02-26 outputs/documents/paper.Rmd
Updates to body of text and tables

---
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    toc: false
    fig_caption: true
    latex_engine: xelatex
    template: templates/svm-latex-ms.tex
bibliography: "references.bib"
header-includes:
  -  \usepackage{hyperref}
  -  \usepackage{amsmath}
biblio-style: apalike
title: "Detecting Hate Speech with GPT-3"
thanks: "Code and data are available at: https://github.com/kelichiu/GPT3-hate-speech-detection. We gratefully acknowledge the support of Gillian Hadfield, the Schwartz Reisman Institute for Technology and Society, and OpenAI for providing access to GPT-3 under the academic access program. We thank Amy Farrow, Haoluan Chen, John Giorgi, Mauricio Vargas Sepúlveda, Monica Alexander, Noam Kolt, and Tom Davidson for helpful discussions and suggestions. Please note that we have added asterisks to racial slurs and other offensive content in this paper, however the inputs and outputs did not have these.  Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca."
author:
- name: Ke-Li Chiu
  affiliation: University of Toronto
- name: Rohan Alexander
  affiliation: University of Toronto and Schwartz Reisman Institute
abstract: "Sophisticated language models such as OpenAI's GPT-3 can generate hateful text that targets marginalized groups. Given this capacity, we are interested in whether large language models can be used to identify hate speech and classify text as sexist or racist? We use GPT-3 to identify sexist and racist text passages with zero-, one-, and few-shot learning. We find that with zero- and one-shot learning, GPT-3 can identify sexist or racist text with an accuracy between 48 per cent and 69 per cent. With few-shot learning and an instruction included in the prompt, the model's accuracy can be as high as 78 per cent. We conclude that large language models have a role to play in hate speech detection, and that with further development language models could be used to counter hate speech and even self-police."
keywords: "GPT-3; natural language processing; quantitative analysis; hate speech."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
endnote: no
graphics: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), echo = FALSE, warning = FALSE)
# library(dplyr)
library(tidyverse)
library(kableExtra)
# library(tidyr)
# library(readr)
```

@@

Natural language processing (NLP) models focus on using normal words, often written text, as their data. For instance, one might have content from a large number of books and want to group them into themes. Sophisticated NLP models are being increasingly embedded in society. For instance, Google Search uses an NLP model called BERT to better understand what is meant by a word given its context. Some sophisticated natural language processing (NLP) models, such as OpenAI's GPT-3, can additionally produce text as an output.

@@

While these, and other, issues give us pause, the dual-use problem, which is that the same technology can be applied to both good and bad uses, provides motivation. For instance, while stylometric analysis can reveal the identity of political dissenters, it can also solve the unknown authorship of historic text [@hovy2016social]. In this paper we are interested in whether given large language models can produce harmful language, can they also identify, or learn to identify, harmful language?

Even though large NLP models do not have a real understanding of language, the vocabularies and the construction patterns of harmful languages can be thought of as known to them. We show that this knowledge can be used to identify abusive language and even hate speech. In particular we consider 120 different extracts that have been categorized as 'racist', 'sexist', or 'neither' in the single-category settings and 243 different extracts in the mixed-category settings. We ask GPT-3 to classify these based on zero-, one-, and few-shot learning, with and without instruction. We find that the model performs best with few-shot learning when an instruction is included. In that setting the model can accurately classify around 78 per cent of the extracts. If language models can be used to identify abusive language, then not only is there potential for them to counter the production of abusive language by humans, but they could also potentially self-police.

@@

<!-- Insert brief description of GPT-3 - the general usage of the model, its transformer-based architecture, what are the inputs-outputs, etc. -->

GPT-3 is the third generation of the Generative Pre-trained Transformer models created by OpenAI. OpenAI is a private company in California that develops artificial intelligence models. Until January 11, when @fedus2021switch announced a Transformer model with a trillion parameters, GPT-3 was the largest, publicly-known, Transformer language model. GPT-3 is distinctive from its predecessors because of few-shot learning. This means that GPT-3 can 'learn' to perform a new task based on only a few examples, expressed in natural language, instead of a fine-tuning process that can require a large amount of data. GPT-3 has led to unexpected NLP applications, such as computational code generation given natural language prompts.

@@

There has been a large amount of research focused on detecting hate speech. And as part of this various hate speech datasets have been created and examined. For instance, @waseem2016hateful create a dataset that captures hate speech in the form of racist and sexist language that includes domain expert annotation. The use Twitter data, and annotate 16,914 tweets: 3,383 as sexist, 1,972 as racist, and 11,559 as neither. They had a high degree of annotator agreement, and most of the disagreements were to do with sexism, and often explained by an annotator lacking apparent context. And, @davidson2017automated trains a classifier to distinguish between hate speech and offensive language. To define hate speech, they use online 'hate speech lexicon containing words and phrases identified by internet users as hate speech'. It is important to note that even these datasets have bias. For instance, @davidson2019racial found racial bias in five different sets of Twitter data annotated for hate speech and abusive language. They found that tweets written in African American English are more likely to be labeled as abusive.

@@

We examine the ability of GPT-3 to identity hate speech in zero-shot, one-shot, and few-shot settings. There are a variety of parameters, such as temperature, that control the degree of text variation. Temperature is a hyper-parameter between zero and one. Lower temperatures mean that the model places more weight on higher-probability tokens. To enhance consistency, the temperature is set to zero in our experiments and not varied. There are two categories of hate speech that are of interest in this paper. The first targets the race of the recipient, and the other targets the gender of the recipient. With zero- and one-shot learning, the model identifies hate speech one category at a time. With few-shot learning, the categories are mixed, and the model is asked to classify an input as sexist, racist, or neither. Zero-shot learning means an example is not provided in the prompt. One-shot learning means that one example is provided, and few-shot means that two or more examples are provided.

@@

For the purposes of one-shot and single-category few-shot learning, the dataset was split in half such that 30 racist comments, 30 sexist comments, and 60 neutral comments were used as inputs and the remainder were used as examples. This division was kept consistent for zero-shot learning to allow for the same set of 120 comments to be classified by all three approaches. With the introduction of 10 static example sets for multi-category few-shot learning, 240 comments were classified for each example set.

@@

With one-shot learning, similarly to the zero-shot learning case, the model is asked to identify one passage of text in one category at a time. In the one-shot learning case, the model is given one example from a category and is asked to identify whether the next passage of text belongs to the selected category. Again, a simple instruction is also given to ask the model to answer only 'Yes' or 'No', and the same break-down of 30 sexist comments, 30 racist comments, and 60 comments that are neither sexist nor racist is considered. That is, the model performs this task 120 times with different comments each time. An example of one-shot learning for recognizing a comment that is labeled as sexist is:

@@

Under the few-shot setting, three examples are provided to the model. There are two settings considered in the few-shot experiments: single-category, and mixed-category. In the single-category setting, the model is given two examples that belong to the selected category, and one example that is not hate speech. The answer is expected to be binary. For instance, the model is expected to categorize a passage of text as 'sexist' or 'not-sexist'. An example of this is:

@@

@@

# Result

@@

The results of the zero-shot experiments are presented in Table \@ref(tab:zeroshot-racism), Table \@ref(tab:zeroshot-sexism), Table \@ref(tab:zeroshot-hate), and \@ref(tab:zeroshot-summary). We assess GPT-3's performance in all settings using accuracy, precision, recall, and F1 score. Accuracy is the proportion of correctly classified comments (hate speech and non-hate speech) out of all comments classified. Precision is the proportion of hate speech comments correctly identified out of all comments identified as hate speech (both correctly and incorrectly). Recall is the number of hate speech comments correctly identified out of all hate speech comments in the data set (both correctly and incorrectly identified). F1 score is the harmonic mean of precision and recall. In the case of hate speech identification it is desirable to have a model with high recall, meaning a model that can identify a relatively high proportion of the hate speech text within a data set, however the F1 score can provide a more well-rounded metric for model performance and comparison.

The model has 34 matches (true positives and negatives) and 26 mismatches (false positives and negatives) in the sexist category, and 41 matches and 19 mismatches in the racist category. The model performs with higher accuracy and precision when identifying racist comments compared with identifying sexist comments. The overall ratio of matches and mismatches is 75:45. In other words, the accuracy in identifying hate speech in the zero-shot setting is 62.5 per cent.

@@

The results of the one-shot learning experiments are presented in Table \@ref(tab:oneshot-racism), Table \@ref(tab:oneshot-sexism), Table \@ref(tab:oneshot-hate), and Table \@ref(tab:oneshot-summary). The model has 37 matches and 23 mismatches in the racist category, and 36 matches and 24 mismatches in the sexist category. In contrast with the result generated from zero-shot learning, the model approximately the same in identifying racist comments compared with identifying sexist comments. The general performance in the one-shot setting is worse than in the zero-shot setting, with lower accuracy, recall, and precision overall. The overall ratio of matches and mismatches is 73:47. In other words, the accuracy of identifying hate speech in the one-shot setting is 60.8 per cent.

@@

The results of the single-category, few-shot learning, experiments are presented in Table \@ref(tab:fewshotsingle-racism), Table \@ref(tab:fewshotsingle-sexism), Table \@ref(tab:fewshotsingle-hate), and Table \@ref(tab:fewshotsingle-summary). The model has 41 matches and 19 mismatches in the racist category, and 42 matches and 18 mismatches in the sexist category. Similar to the one-shot setting, the model performs almost equally well at identifying racist comments compared with identifying sexist comments. The general performance in the few-shot learning setting is similar to the performance in the one-shot learning setting as well. The overall ratio of matches and mismatches is 83:37, or around 69.2 per cent, with higher precision and lower recall compared with the one-shot learning setting.

@@

knitr::kable(table,
             caption="Performance of mixed-category few-shot learning in racist text classification",
               # col.names=c("", "Percent"),
             booktabs=TRUE,
             linesep = "") %>%
  kableExtra::row_spec(seq(2, 20, by=2), hline_after = TRUE)

@@

knitr::kable(table,
             caption="Performance of mixed-category few-shot learning in racist text classification, with instruction",
               # col.names=c("", "Percent"),
             booktabs=TRUE,
             linesep = "") %>%
  kableExtra::row_spec(seq(2, 20, by=2), hline_after = TRUE)

@@

<!-- Fix below -->
In the mixed-category few-shot setting, different example sets yield different accuracy rates for racist and sexist comments with noticeable improvement over the single-category approaches. Given our definition of true negative in the mixed-category setting (see Section \@ref(sec:few-shot-learning----mixed-category)), large number of true negatives, present in the dataset

Without an instruction included in the prompt, the accuracy rates range from 57 per cent (Example Set 9) to 75 per cent (Example Set 2). Adding an instruction improves the performance, and the highest accuracy rate increases from 75 per cent (Example Set 2) to 78 per cent (Example Set 1). Overall accuracy rates also increase in almost every example set after adding an instruction in the prompt. The large difference in accuracy rates between the example sets suggests that the content of the examples may matter. It may be that deliberately curated examples would enable better accuracy.

>>>>

---
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    toc: false
    fig_caption: true
    latex_engine: xelatex
    template: templates/svm-latex-ms.tex
bibliography: "references.bib"
header-includes:
  -  \usepackage{hyperref}
  -  \usepackage{amsmath}
biblio-style: apalike
title: "Detecting Hate Speech with GPT-3"
thanks: "Code and data are available at: https://github.com/kelichiu/GPT3-hate-speech-detection. We gratefully acknowledge the support of Gillian Hadfield, the Schwartz Reisman Institute for Technology and Society, and OpenAI for providing access to GPT-3 under the academic access program. We thank Amy Farrow, Haoluan Chen, John Giorgi, Mauricio Vargas Sepúlveda, Monica Alexander, Noam Kolt, and Tom Davidson for helpful discussions and suggestions. Please note that we have added asterisks to racial slurs and other offensive content in this paper, however the inputs and outputs did not have these.  Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca."
author:
- name: Ke-Li Chiu
  affiliation: University of Toronto
- name: Rohan Alexander
  affiliation: University of Toronto and Schwartz Reisman Institute
abstract: "Sophisticated language models such as OpenAI's GPT-3 can generate hateful text that targets marginalized groups. Given this capacity, we are interested in whether large language models can be used to identify hate speech and classify text as sexist or racist. We use GPT-3 to identify sexist and racist text passages with zero-, one-, and few-shot learning. We find that with zero- and one-shot learning, GPT-3 can identify sexist or racist text with an accuracy between 56.7 per cent and 68.3 per cent depending on the category of text and type of learning. With few-shot learning, the model's accuracy can be as high as 88.3 percent. We conclude that large language models have a role to play in hate speech detection, and that with further development language models could be used to counter hate speech and even self-police."
keywords: "GPT-3; natural language processing; quantitative analysis; hate speech."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
endnote: no
graphics: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), echo = FALSE, warning = FALSE)
# library(dplyr)
library(tidyverse)
library(kableExtra)
# library(tidyr)
# library(readr)
```

@@

Natural language processing (NLP) models focus on using normal words, often written text, as their data. For instance, one might have content from a large number of books and want to group them into themes. Sophisticated NLP models are being increasingly embedded in society. For instance, Google Search uses an NLP model called BERT to better understand what is meant by a word given its context. Some sophisticated natural language processing (NLP) models, such as OpenAI's Generative Pre-trained Transformer 3 (GPT-3), can additionally produce text as an output.

@@

While these issues (among others) give us pause, the dual-use problem, which is that the same technology can be applied to both good and bad uses, provides motivation. For instance, while stylometric analysis can reveal the identity of political dissenters, it can also solve the unknown authorship of historic text [@hovy2016social]. In this paper we are interested in whether large language models, given that they can produce harmful language, can also identify, or learn to identify, harmful language.

Even though large NLP models do not have a real understanding of language, the vocabularies and the construction patterns of harmful languages can be thought of as known to them. We show that this knowledge can be used to identify abusive language and even hate speech. In particular we consider 120 different extracts that have been categorized as 'racist', 'sexist', or 'neither' in the single-category settings (zero-shot, one-shot, and few-shot) and 243 different extracts in the mixed-category few-shot settings. We ask GPT-3 to classify these based on zero-, one-, and few-shot learning, with and without instruction. We find that the model performs best with mixed-category few-shot learning. In that setting the model can accurately classify around 85.6 percent of the racist extracts and 88.3 percent of sexist extracts. If language models can be used to identify abusive language, then not only is there potential for them to counter the production of abusive language by humans, but they could also potentially self-police.

@@

<!-- Insert brief description of GPT-3 - the general usage of the model, its transformer-based architecture, what are the inputs-outputs, etc. - ADD MORE? -->

GPT-3 is the third generation of the Generative Pre-trained Transformer models created by OpenAI, a private company in California that develops artificial intelligence models. GPT-3 is an autoregressive NLP model that can perform a variety of tasks, including responding to questions, summarizing and parsing text, translation, and classification. Interactions with the model involve inputting some text as a prompt and GPT-3 returning a text completion according to that prompt.

Until January 11, when @fedus2021switch announced a Transformer model with a trillion parameters, GPT-3 was the largest, publicly-known, Transformer language model. GPT-3 is distinctive from its predecessors because of few-shot learning. This means that GPT-3 can 'learn' to perform a new task based on only a few examples, expressed in natural language, instead of a fine-tuning process that can require a large amount of data. GPT-3 has led to unexpected NLP applications, such as computational code generation given natural language prompts.

@@

There has been a large amount of research focused on detecting hate speech. As part of this process, various hate speech datasets have been created and examined. For instance, @waseem2016hateful create a dataset that captures hate speech in the form of racist and sexist language that includes domain expert annotation. The use Twitter data, and annotate 16,914 tweets: 3,383 as sexist, 1,972 as racist, and 11,559 as neither. They had a high degree of annotator agreement, and most of the disagreements were to do with sexism, and often explained by an annotator lacking apparent context. And, @davidson2017automated trains a classifier to distinguish between hate speech and offensive language. To define hate speech, they use online 'hate speech lexicon containing words and phrases identified by internet users as hate speech'. It is important to note that even these datasets have bias. For instance, @davidson2019racial found racial bias in five different sets of Twitter data annotated for hate speech and abusive language. They found that tweets written in African American English are more likely to be labeled as abusive.

@@

We examine the ability of GPT-3 to identify hate speech in zero-shot, one-shot, and few-shot settings. There are a variety of parameters, such as temperature, that control the degree of text variation. Temperature is a hyper-parameter between zero and one. Lower temperatures mean that the model places more weight on higher-probability tokens. To enhance consistency, the temperature is set to zero in our experiments and not varied. There are two categories of hate speech that are of interest in this paper. The first targets the race of the recipient, and the other targets the gender of the recipient. With zero-, one-, and few-shot single-category learning, the model identifies hate speech one category at a time. With few-shot mixed-category learning, the categories are mixed, and the model is asked to classify an input as sexist, racist, or neither. Zero-shot learning means an example is not provided in the prompt. One-shot learning means that one example is provided, and few-shot means that two or more examples are provided.

@@

For the purposes of one-shot and single-category few-shot learning, the dataset was split in half such that 30 racist comments, 30 sexist comments, and 60 neutral comments were used as inputs and the remainder were used as examples. This division was kept consistent for zero-shot learning to allow for the same set of 120 comments to be classified by all three approaches. With the introduction of 10 static example sets for multi-category few-shot learning, 240 comments were classified for each example set. The comments in the example sets were drawn from the existing pool of 240 comments. For each example set, the comment from each category being used in the example set was dropped from the input dataset and an additional comment from each original dataset (one 'racist', one 'sexist', and one 'neither') was added in as a replacement. This replacement was kept consistent each time, for a total of 243 unique comments classified in the few-shot mixed-category setting.

@@

With one-shot learning, similarly to the zero-shot learning case, the model is asked to identify one passage of text in one category at a time. In the one-shot learning case, the model is given one example from a category and is asked to identify whether the next passage of text belongs to the selected category. Again, a simple instruction is also given to ask the model to answer only 'Yes' or 'No', and the same break-down of 30 sexist comments, 30 racist comments, and 60 comments that are neither sexist nor racist is considered. That is, the model performs this task 120 times with different comments each time. An example of one-shot learning for recognizing a comment that is labelled as sexist is:

@@

Under the few-shot setting, three examples are provided to the model. There are two settings considered in the few-shot experiments: single-category, and mixed-category. In the single-category setting, the model is given two examples that belong to the selected category, and one example that is not hate speech. The answer is expected to be binary. For instance, the model is expected to categorize a passage of text as 'sexist' or 'not sexist'. An example of this is:

@@

<!-- Is the description of how the example sets were selected too detailed here? -->

@@

# Results

We assess GPT-3's performance in all settings using accuracy, precision, recall, and F1 score. Accuracy is the proportion of correctly classified comments (hate speech and non-hate speech) out of all comments classified. Precision is the proportion of hate speech comments correctly identified out of all comments identified as hate speech (both correctly and incorrectly). Recall is the number of hate speech comments correctly identified out of all hate speech comments in the data set (both correctly and incorrectly identified). F1 score is the harmonic mean of precision and recall. In the case of hate speech identification it is desirable to have a model with high recall, meaning a model that can identify a relatively high proportion of the hate speech text within a data set. However, the F1 score can provide a more well-rounded metric for model performance and comparison.

@@

The results of the zero-shot experiments are presented in Tables \@ref(tab:zeroshot-racism), \@ref(tab:zeroshot-sexism), \@ref(tab:zeroshot-hate), and \@ref(tab:zeroshot-summary). The model has 34 matches (true positives and negatives) and 26 mismatches (false positives and negatives) in the sexist category, and 41 matches and 19 mismatches in the racist category. The model performs with higher accuracy and precision when identifying racist comments compared with identifying sexist comments. The overall ratio of matches and mismatches is 75:45. In other words, the accuracy in identifying hate speech in the zero-shot setting is 62.5 per cent.

@@

The results of the one-shot learning experiments are presented in Tables \@ref(tab:oneshot-racism), \@ref(tab:oneshot-sexism), \@ref(tab:oneshot-hate), and \@ref(tab:oneshot-summary). The model has 37 matches and 23 mismatches in the racist category, and 36 matches and 24 mismatches in the sexist category. In contrast with the result generated from zero-shot learning, the model approximately the same in identifying racist comments compared with identifying sexist comments. The general performance in the one-shot setting is worse than in the zero-shot setting, with lower accuracy, recall, and precision overall. The overall ratio of matches and mismatches is 73:47. In other words, the accuracy of identifying hate speech in the one-shot setting is 60.8 per cent.

@@

The results of the single-category, few-shot learning, experiments are presented in Tables \@ref(tab:fewshotsingle-racism), \@ref(tab:fewshotsingle-sexism), \@ref(tab:fewshotsingle-hate), and \@ref(tab:fewshotsingle-summary). The model has 41 matches and 19 mismatches in the racist category, and 42 matches and 18 mismatches in the sexist category. Similar to the one-shot setting, the model performs almost equally well at identifying racist comments compared with identifying sexist comments. The general performance in the few-shot learning setting is similar to the performance in the one-shot learning setting as well. The overall ratio of matches and mismatches is 83:37, or around 69.2 per cent, with higher precision and lower recall compared with the one-shot learning setting.

@@

knitr::kable(table,
             caption="Performance of mixed-category few-shot learning in text classification",
               # col.names=c("", "Percent"),
             booktabs=TRUE,
             linesep = "") %>%
  kableExtra::row_spec(seq(2, 20, by=2), hline_after = TRUE)

@@

knitr::kable(table,
             caption="Performance of mixed-category few-shot learning in text classification, with instruction",
               # col.names=c("", "Percent"),
             booktabs=TRUE,
             linesep = "") %>%
  kableExtra::row_spec(seq(2, 20, by=2), hline_after = TRUE)

@@

In the mixed-category few-shot setting, different example sets yield different accuracy rates for racist and sexist comments with noticeable improvement over the single-category approaches. Our definition of true negative in the mixed-category few-shot setting (see Section \@ref(few-shot-learning----mixed-category)) resulted in a large number of true negatives in our analysis, thus recall and F1 score may be better metrics for measuring model performance in this context. Mixed-category few-shot learning without instruction had noticeably better F1 scores for both racist and sexist comments than either zero-shot or one-shot learning. With instruction added, mixed-category few-shot learning performed similarly well for racist text identification, however the model performed relatively poorly in identifying sexist speech, with an F1 score of 62.97 percent overall and a recall of 50.17 percent. Overall, it appears as though GPT-3 is most effective at identifying both racist and sexist comments in the mixed-category few-shot learning setting, without instruction.

Examining number of classification matches for each example set, the overall proportion of matches increases in almost every example set after adding an instruction in the prompt. The differences in accuracy, precision, and recall between the example sets suggests that the content of the examples may matter as well. It may be that deliberately curated examples would enable better accuracy. Adding an instruction to the prompt is useful for restricting the output of the model to more clear-cut classifications. Depending on the purposes of classification, it appears as though adding an instruction can improve model performance by some metrics, however this may vary by the category of text considered and still produce the low recall observed in many of the example sets classified with instruction.
